Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-12-15 Manuscript/Manuscript.tex
Edit missing data matrix

\Abstract{
Tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
Including methods for ignorable and non-ignorable missingness.
}

@@

We talk of multilevel data when there is some kind of hierarchy or
clustering in a dataset. In the typical case, individuals are nested
within groups, but there are many different types of multilevel data. In
the medical field, clustering occurs at e.g., the hospitals/center level
in registry data, or at the study-level in meta-analyses (IPDMA). In the
social sciences and official statistics we can find clustering e.g.~at
the country-level, or as imposed by the sampling design. In this paper,
we will refer to the grouping variable as `cluster', and the grouped
variable as `(sample) unit'.\footnote{Alternatives to unit:
  participant/response/record.} For reasons of brevity, we only discuss
clustering between units, not within units (such as in timeseries or
longitudinal data).\footnote{Also, add that we'll only discuss two
  levels, no more?}

Analyzing multilevel data requires special care, compared to `regular',
single level data. The cluster to which a unit belongs may influence the
unit-level observations, and since clusters are made up of units,
clusters depend on units as well \citep{hox17}. These relations can and
should be taken into account when developing analysis models for
multilevel data.\footnote{Explain ICC here? The percentage of variance
  attributed to the cluster-level is expressed by the intra-class
  coefficient (ICC). The ICC can also be interpreted as the expected
  correlation between two randomly sampled units in same cluster. So if
  the ICC is high, a lot of variability in a variable is due to the
  clustering, which should be modeled accordingly.} Multilevel models
typically include separate intercepts for each cluster, which relieves
one restriction imposed by single-level models: equal group means across
clusters. Additionally, there may be random predictor effects and/or
random error terms (residual error variances), see e.g. \citet{hox17}
and \citet{jong21}.\footnote{Add that heterogeneity refers to
  variability within clusters.} There are many names for models that
take clustering into account. Some popular examples are `multilevel
models', `hierarchical models', `mixed effect models' and `random effect
models'.

@@

@@

Multilevel data is not spared the ubiquitous problem of missing
information. Just as in single level data, missingness may occur at the
unit level. But with multiple levels of data comes the potential for
missingness at multiple levels. Missingness in multilevel data can be
categorized into two general patterns: systematical missingness and
sporadic missingness, see \citet{resc13}. In figure 1, we show a dataset
with units in the rows and variables in the columns, there are 5 units
nested within 2 clusters, and 3 variables of interest. Variable
\texttt{X1} is completely observed. Variable \texttt{X2} is
systematically missing, \texttt{X3} is sporadically missing.\footnote{Explain
  why.} Systematic missingness can be further subdivided into unobserved
constants (same value per cluster) and non-measured random variables
(which may differ per unit within clusters). In Figure 1, the former
implies that the unobserved values for units 4 and 5 on variable
\texttt{X2} would be equal. With the latter, the values may differ.
Depending on the missing data pattern, there are more or less optimal
way of accommodating the missingness.

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations. For reasons of brevity, we focus on
imputation by chained equations, although JOMO is available in
\pkg{mice} as well. Other useful packages for incomplete multilevel data
include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.\footnote{Rephrase:
  Some level of knowledge on multilevel models is assumed. We're
  providing an overview of implementations. It's up-to the reader to
  decide which multilevel strategy suits their data. So we won't go into
  detail for the different methods (and equations). Refer to
  \citet{meng94}, \citet{audi18}, and \citet{grun18}. This paper is just
  a software tutorial. We'll keep it practical.} We introduce three case
studies to illustrate the workflow.

@@

For each case study we focus on a different aspect to illustrate how to
impute incomplete multilevel data. In the \texttt{mice::popmis} data, we
show the advantages of including the multilevel structure of the data
into the imputation model. In the \texttt{metamisc::impact} data we'll
show how to induce missingness and solve it in real-world data. In the
\texttt{GJRM::hiv} we provide novel methodology\footnote{not really, the
  methods exist already, but how to show that this is something new and
  exciting?} for imputing MNAR missingness according to the Heckman
model.

For each case study we'll look at least at: 1) the incomplete data; 2)
the imputation model; 3) the imputed data; and 4) how to obtain pooled
estimates for the analysis of scientific interest.

@@

The missingness in this dataset is induced conform MAR and MNAR
mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows
the systematic nature of the missingness.

>>>>

\Abstract{
Multilevel data is not spared the ubiquitous problem of missing
information. This is a tutorial paper on imputing incomplete multilevel
data with \pkg{mice}. Including methods for ignorable and non-ignorable
missingness. Footnotes in the current version show work in
progress/under construction. The last section is not part of the
manuscript, but purely for reminders.
}

@@

Research into any field with a hierarchical or clustered nature of
observations may yield multilevel data. In the typical case, individuals
are nested within groups, but there are many different types of
multilevel data. In the medical field, clustering occurs at e.g., the
hospital or center level in registry data, or at the study-level in
meta-analyses (IPDMA\footnote{remove or write in full?}). In the social
sciences and official statistics we can find clustering e.g.~at the
country-level, or as imposed by a multi-stage sampling design. For the
sake of legibility, we will refer to the grouping variable as `cluster',
and the grouped variable as `(sample) unit' throughout this
paper.\footnote{Alternatives to unit: participant/response/record.} And,
for reasons of brevity, we only discuss clustering between units, not
within units (e.g., timeseries or longitudinal data).\footnote{Also, add
  that we'll only discuss two levels, not more?}

Analyzing multilevel data requires special care, compared to `regular',
single level data. For instance, there may be cross-level interactions
between unit-level variables and cluster-level variables. The cluster to
which a unit belongs may influence the unit-level observations, and vice
versa for the the units that make up the cluster \citep{hox17}. These
relations can and should be taken into account when developing analysis
models for multilevel data.\footnote{Explain ICC here? The percentage of
  variance attributed to the cluster-level is expressed by the
  intra-class coefficient (ICC). The ICC can also be interpreted as the
  expected correlation between two randomly sampled units in same
  cluster. So if the ICC is high, a lot of variability in a variable is
  due to the clustering, which should be modeled accordingly.}
Multilevel models typically include separate intercepts for each
cluster, which relieves one restriction imposed by single-level models:
equal group means across clusters. Additionally, there may be random
predictor effects and/or random error terms (residual error variances),
see e.g. \citet{hox17} and \citet{jong21}.\footnote{Add that
  heterogeneity refers to variability within clusters.} There are many
names for models that take clustering into account. Some popular
examples are `multilevel models', `hierarchical models', `mixed effect
models' and `random effect models'.

@@

Multilevel data is not spared the ubiquitous problem of missing
information. Just as in single level data, missingness may occur at the
unit level. But with multiple levels of data comes the potential for
missingness at multiple levels. Missingness in multilevel data can be
categorized into two general patterns: systematical missingness and
sporadic missingness, see \citet{resc13}. In figure 1, we show a dataset
with units in the rows and variables in the columns, there are 5 units
nested within 3 clusters, and 3 variables of interest. Variable
\texttt{X1} is completely observed. Variable \texttt{X2} is
systematically missing, \texttt{X3} is sporadically missing.\footnote{Explain
  why.} Systematic missingness can be further subdivided into unobserved
constants (i.e., the same value within clusters) and non-measured random
variables (which may differ per unit within clusters). In Figure 1, the
former implies that the unobserved values for units 3 and 4 on variable
\texttt{X2} would be equal. With the latter, the values may differ.
Depending on the missing data pattern\footnote{add missing data
  mechanisms here?}, there are more or less optimal way of accommodating
the missingness.

@@

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations. For reasons of brevity, we focus on
imputation by chained equations\footnote{add that JOMO is available in
  \pkg{mice} as well?}. Other useful packages for incomplete multilevel
data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.\footnote{Rephrase:
  Some level of knowledge on multilevel models is assumed. We're
  providing an overview of implementations. It's up-to the reader to
  decide which multilevel strategy suits their data. So we won't go into
  detail for the different methods (and equations). Refer to
  \citet{meng94}, \citet{audi18}, and \citet{grun18}. This paper is just
  a software tutorial. We'll keep it practical.}

@@

For each case study we focus on a different aspect to illustrate how to
impute incomplete multilevel data. In the \texttt{mice::popmis} data, we
show the advantages of including the multilevel structure of the data
into the imputation model. In the \texttt{metamisc::impact} data we'll
show how to induce missingness and solve it in real-world data. In the
\texttt{GJRM::hiv} we provide novel methodology\footnote{not really, the
  methods exist already, but how to show that this is something new and
  exciting?} for imputing MNAR missingness according to the Heckman
model. For all case studies we discuss the nature of the incomplete
data, the imputation model(s), and evaluation of the imputed data.

@@

The missingness in this dataset is induced conform MAR and MNAR
mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows
that just one variable is incomplete \textbf{{[}the next part is not yet
updated to reflect this{]}}.
